---
title: "Final Report: Life After Death- The Prevailing Microbial Species of Recently-Deceased Individuals"
author: "Alfredo Hans Locht"
date: "November 29, 2017"
output:
    github_document: default
    word_document: default
bibliography: references.bib
csl: bioinformatics.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("png")
load("output/Thanatomicrobiome_phyloseq.RData")

```

# Introduction

Add about 2-3 pages here. Across the whole manuscript, you should cite at least 20 peer reviewed articles.

# Methods

## Study design

[@Hyde2013] obtained cadavers donated to the Southeast Texas Applied Forensic Science (STAFS) Facility at the Center for Biological Field Studies (CBFS) at Sam Houston State University (SHSU). Both CBFS and STAFS are located in a humid, subtropical climate characterized by a large distribution of pine trees and acidic soils in an area 5 km North of Huntsville, Texas. STAFS is a sectioned-off, two-acre area within the larger, 243 acre CBFS complex which is covered with pine trees and herbaceous plants. The STAFS facility is considered to be a “willed donations” center, so donors or family members of the deceased are required to sign a form in which they indicate that they understand that their donated corpse will be used for scientific research and teaching purposes. 

This form also included information regarding the methods that would be used to study the cadavers. Specifically, the study informed the donors and/or next of kin that the body would be left outside to decompose in natural and aerobic conditions, and that samples pertaining to different types of research (bacterial, tissue, and entomological) would be acquired at different stages of the decomposition to support the associated research. Upon request, the family members could also choose to be debriefed about the scientific knowledge or educational project that was achieved during the course of the study and as a result of the donated cadaver. 

## Sample Origin and Sequencing

### Cadaver Demographics

The two cadavers donated to the center that were studied, STAFS 2011-016 and STAFS 2011-006, were both males of ages 68 and 52 respectively, both of Caucasian descent, were corpses that had not been autopsied, and were considered to be in the early stages of decomposition according to a qualitative body-scoring index developed by [@Megyesi2005]. This index relied on visual observation of the cadavers for signs of the bloat process [@Hyde2013]. STAFS 2011-016 had a previous medical history relating to the cause of death (acute myocardial infarction), been frozen for 89 days, and had been subsequently defrosted in a fridge cooler for 143 days before being placed outside to decompose. STAFS 2011-006 did not have any pertaining medical history other than the cause of death (carbon monoxide poisoning), had been frozen for 13 days, and had subsequently defrosted for 6 days in a fridge cooler before decomposition placement. The time since death was unknown for both cadavers. 

### Sampling Methods

Bacterial sampling of each cadaver was obtained from multiple body sites at two different time points, Placement (Pre-Bloat Phase) and Conclusion (End-Bloat Phase). The Pre-Bloat samples were taken right before the corpse was placed in a supine position outside with no clothing to decompose in natural conditions, while the End-Bloat samples required that the cadaver be dissected in order to gain access to the internal organs. Because this procedure was intrusive and compromised the structural integrity of the cadaver, it marked the end time-point of the sampling. 

Pre-bloat samples were taken as either swabs or “scrapes” from the mouth and rectum by swabbing/scraping with a sterile cotton applicator which was subsequently cut into a sterile PBS solution and prepped for sequencing using the PowerSoil DNA Isolation Kit (MoBio). Bloat samples were taken from the corpses at a time which they were identified to have purged their bodily fluids according to the body scoring system [@Megyesi2005]. Maggots found on the cadavers were also removed at this point with sterile plastic spatulas for sampling purposes. 

The samples obtained from different parts of the body or organs were sampled whenever those body parts or organs still existed. Specifically, STAFs 2011-016 no longer retained any evidence of a sigmoidal colon at the end-bloat sampling.

### Limting Disclaimers

Conditions surrounding the death of any given cadaver can have high variability in confounding factors which can directly influence the rate and pattern of decomposition [@Bates2016; @Cockle2015; @Ferreira2013; @Hauther2015; @Hyde2013; @Vass2011]. Intrinsic factors such as cause of death, physical trauma, and postmortem care can directly interact with a number of other, extrinsic factors such as temperature and humidity of environment in manners which have not yet been efficiently characterized [@Cockle2015]. [@Hyde2013] forewarn that because of the inherent inconsistency of the STAFS willed donations program (cadavers can be obtained on a random or unexpected basis) and the associated complications in conducting decompositional studies on cadavers, data acquired for both cadavers cannot be compared directly but rather, should act as a complimentary set of data. Although the authors sought to control or reduce the amount of confounding variables to the best of their knowledge, they were able to identify some of the limiting qualities of their study. These disclaimers are investigated below: 

In order to establish a baseline of decomposition of both cadavers, they were initially qualified by a visual scoring system developed by [@Megyesi2005]. A summary of the system includes visible cues that might indicate at what step of the decomposition process a particular cadaver may be at. Using this qualitative method, both cadavers were considered to be in the early stages of decomposition as no bloat had yet occurred. This system was also used to identify the end-point of sampling when the purging of bodily fluids occurs; however, due to its qualitative nature, the cadavers did not spend an equal amount of time decomposing outside. 

STAFS 2011-006 spent seven days out in September while STAFS 2011-016 spent fourteen days out in November. Because the body scoring system developed by [@Megyesi2005] is completely dependent on temperature, [@Hyde2013] retroactively calculated the heat-energy units for each cadaver which are a function of Accumulated Degree-Days (ADD), an arbitrary calculation of how much thermal energy a sample is exposed to dependent on ambient temperature and a chosen starting point. The results of the calculations were 197.14 heat-energy units for STAFS 2011-006 and 199.97 heat-energy units for STAFS 2011-016 when a thermal minimum of 4 degrees Celsius was chosen. Because November temperatures in Texas are less than September temperatures, it is logical to see that STAFS 2011-016 would probably have to spend twice as long outside in November to achieve the same level of decomposition as STAFS 2011-006 in September, but again, this is based on a visual and qualitative scoring index. Furthermore, the authors mention that choosing a different thermal minimum would result in differing calculations from the ones above [@Hyde2013].

Other variables speaking towards the physical characteristics such as height and weight of each male cadaver were unknown, but estimated to be of average height and above-average weight in comparison to an average American male [@Hyde2013].

## Computational

These are the methods you used to do your bioinformatic analyses analyses. Should probably be between 0.5 and 1 pages. At a very minimum should include citations for DADA2 and phyloseq if you are doing an amplicon study, or other citations as appropriate.

# Results

```{r}
plot_richness(phyloseq_obj,
              x = "collection_date_s",
              measures = c("Shannon", "Simpson"),
              color = "sample_description_s") +
  xlab("Source of Sample") +
  geom_jitter(width = 0.2) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))
  theme_bw()
```

```{r}
plot_richness(phyloseq_obj,
              x = "sample_description_s",
              measures = c("Shannon", "Simpson"),
              color = "collection_date_s") +
  xlab("Source of Sample") +
  geom_jitter(width = 0.2) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))
  theme_bw()
```

```{r}
gp_ch <- subset_taxa(phyloseq_obj, Phylum == "Firmicutes")
plot_bar(gp_ch,
  fill = "Family",
  title = "Firmicutes Phyla Acrosss Distribution Across Samples")
```

```{r}
plot_richness(gp_ch,
              x = "sample_description_s",
              measures = c("Shannon", "Simpson"),
              color = "collection_date_s") +
  xlab("Source of Sample") +
  geom_jitter(width = 0.2) +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1))
  theme_bw()
```





```{r}
gp_ch2 <- subset_taxa(phyloseq_obj, Phylum == "Actinobacteria")
plot_bar(gp_ch2,
  fill = "Family",
  title = "Actinobacteria Phyla Acrosss Distribution Across Samples")
```


```{r}
gp_ch3 <- subset_taxa(phyloseq_obj, Class = "Chlostridia")
plot_bar(gp_ch3,
  fill = "Order",
  title = "Firmicutes Phyla Acrosss Distribution Across Samples")
```

```{r}
gpac <- subset_taxa(phyloseq_obj, Phylum == "Firmicutes")
gpac <- prune_taxa(names(sort(taxa_sums(gpac),TRUE)[1:300]), gpac)
plot_heatmap(phyloseq_obj,
             method = "NMDS"
             )









```











## Subsections are ok in the results section too

```{r your-code-here}
# Add code chunks as needed for your analyses
# For most analyses, I would recommend splitting the
# intensive computational part into a seperate R script file
# and then just load your libraries and the data object here
# using the `load()` function. Ask me for clarification if this is
# unclear.
```

In addition to a minimum of 5-10 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion. 2-3 pages.

# Discussion

Add around 3-4 pages interpreting your results and considering future directions one might take in analyzing these data.

# Sources Cited
